A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-22 paint-contour-tree.tex
New figures

\title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
\date{}

@@

\begin{abstract}
Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, there has been significant previous work on algorithms for their efficient computation.

Some of these previous results are describe as optimal, however they are only worst case optimal.  Here we present an output sensitive approach to computing contour trees.  Our approach leads to a faster algorithm when the contour tree is nicely behaved (i.e. more balanced), yet is still optimal in the worst case.  Our approach is novel and we believe will lead to faster algorithms for other applications as well.
\end{abstract}

\section{Introduction}
Describe what reeb graphs and contour trees are.

\medskip
\noindent
Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.

\medskip
\noindent
Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.

@@

\paragraph{Contour Trees.}
As the application areas of contour trees usually involve very large data sets, highly efficient algorithms are not only desirable but often necessary.  As such there have been a long sequence of papers with improved running times for computing contour trees.  Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh. Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making several passes over the input.

Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, where $n$ is the number of mesh vertices.  Their algorithm works in 3 simple stages.  The first stage consist of a top down sweep to compute what is known as the merge (or join) tree, which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  Their main contribution is the third and final step where 
they present an $O(n)$ time algorithm to merge together the merge and split trees to get the contour tree.  Chiang \etal \cite{cllr-sooscctmp-05} 
have since

%The most expensive part of the Carr \etal algorithm comes from sorting.  They sort all vertices of the mesh to facilitate an easing sweeping of the mesh.
%asdfas \cite{} improve the running to $O()$ by sorting only critical points of the mesh.  However, without looking at the non-critical points in the mesh 
%one does not know how the critical points connect to each other in the contour tree.  To overcome this \cite{} use ``mountain climbers'' from each saddle 
%to determine which critical points they belong to above.

\paragraph{Output Sensitivity.}
Let $t$ be the number of critical points in the input mesh.  There is room to improve the results of Carr \etal, as $t$ could potentially be much smaller than $n$.  In particular, Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 3 dimensional structured meshes.  Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ which works for structured or unstructured meshes in any dimension.

As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, one can show that any algorithm to compute the contour tree in the worst case takes $\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only worst case optimal.  In this paper we will consider an output sensitive approach which takes into account the shape of the contour tree.

\paragraph{Reeb Graphs.} UNDER CONSTRUCTION.
There have also been a number of result for the more general Reeb graph, i.e. when the loopless assumption is removed.  Shinagawa and Kunii \cite{sk-crgacs-91} provided the first provably correct algorithm for the case of triangulated $2$-manifolds. Cole-McLaughlin \etal \cite{cehnp-lrbm-03} later improved the running time to $O(n\log n)$.

\subsection{Our Approach and Contribution.}

\paragraph{To Sort or not to Sort.}
As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour tree by first constructing the merge and split trees. 
As vertices appear in sorted order in any root to leaf path in either tree, sorting the input vertices is a natural step.  However, not all vertices must be sorted together.  In particular, Chiang \etal \cite{cllr-sooscctmp-05} improved the running time by observing that as the merge and split tree only consists of critical points, it suffices to only sort critical points.  
%
%However, without looking at the non-critical points in the mesh 
%one does not know how the critical points connect to each other in the contour tree.  \cite{} easily overcome this by sending out increasing monotone paths (i.e. ``mountain climbers'') from each critical point to determine which critical points they connect to above.
%
If a large fraction of the vertices appear in a single root to leaf path in the merge or split tree, paying to sort all the critical 
points is unavoidable in order to compute the merge and split trees.  However, if the merge and split trees are balanced then sorting all the critical points is potentially very costly.

There is also a more fundamental problem with the now standard approach of first computing the merge and split trees in order to compute the contour tree.  In particular, this approach can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  This is because it is possible to have merge and split trees which are long and skinny (i.e. costly to compute) whereas the contour tree is balanced.  For example consider the case when the mesh looks like two (thickened) balanced binary trees, one upside down, and joined at their roots.  In this case the contour tree is potentially cheap to compute as there is no long sorted path.  However, the merge and split trees will have long sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), and therefore require $\Omega(n\log n)$ time to compute.

\paragraph{Our Contribution.} UNDER CONSTRUCTION.
Give a high level overview.  Emphasize the spilling partitioning scheme used to avoid sorting, and say it may be useful else where.  State our running time.  Say it may be optimal (we still need to investigate).  It matches previous on worst case (i.e. optimal in worst case).

@@

\subsection{Shrubs, heavy paths, and light paths}

@@

\begin{definition}
 For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
\end{definition}

@@

The following lemma demonstrates that heavy paths can ``pay'' for themselves.

\begin{lemma}
\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
% Then, 
% \begin{compactenum}
%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
% \end{compactenum}
\end{lemma}
\begin{proof} \ignore{
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
}
For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
 If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For light paths, $|p| = O(|H_p|)$.
\end{proof}

\ignore{
For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
Our eventual goal will be to show that by choosing the right path decomposition, this excess 
can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
}

There are some light paths that we can also ``pay" for.  Consider any light path $p$ in the shrub.  We will refer to the 
\emph{heavy support chain} of $p$ as the heavy ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
light path (i.e. it is a chain of paths adjacent to $p$).

\begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the heavy support chain of $p$.
%
%
% there exists
%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
%and all $q_i$'s are heavy.
\end{definition}

Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths
given by $\pmax(T)$.
We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported light paths, 
and the remaining nodes are the unsupported light paths.

@@

We split the summation into heavy, light, and unsupported light paths.
\begin{eqnarray*} 
\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
\end{eqnarray*}
The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
The following claim shows that this in turn is at most the last term.

@@

\begin{proof} 
Pick $p \in \cL'$. 
%Let the first $|H_p|/200$ vertices the lie in heavy paths be denoted as $H'_p$. 
%Observe that for any $v \in H'_p$, 
%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in heavy paths,
%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%
%
As we traverse the heavy support chain of $p$, there
are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
Since all the vertices lie in heavy paths,
we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.

Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
since paths which have an ancestor-descendant relationship have disjoint heavy support chains.  
However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.

@@

The first term (unsupported light paths) can be charged to the second term (supported light paths).
This is where the critical \Lem{cu-root} plays a role.

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
\end{proof}

\ignore{
Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.

\begin{lemma} \label{lem:light} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{lemma}

@@

\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into heavy and light paths.
We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
\end{proof}

>>>>

\title{A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm}
\date{}

@@

\begin{abstract}
Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
there has been significant previous work in developing algorithms for their efficient computation.

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results (on all instances requiring $\Omega(n\alpha(n))$ time).
In particular, our algorithm runs in $O(n\alpha(n))$ time when the tree is balanced, yet still 
matches the previous worst case $O(n\log n)$ bound when the tree is long and skinny.

In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
such tools as heavy-light path decompositions.
\end{abstract}

\section{Introduction}

Let $f$ be a real valued function defined over a manifold embedded in $\RR^d$.  
For any real value $\alpha$, the level set $f^{-1}(\alpha)$ can be decomposed into its connected components, or ``contours''.
The Reeb graph is a fundamental topological structure which represents the evolution of these contours as one sweeps the surface 
from $-\infty$ to $\infty$ according to $f$.  Often this graph is acyclic, in which case it is 
called a \emph{contour tree}.  This is the case if, for example, the input is a surface mesh of a mountain range 
(and the contour tree would encode when mountain peeks form and how they merge together).

There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
Often these applications involve massive data sets, and so consequently there has also been a long series 
of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
As an initial step these algorithms require that all $n$ contour tree vertices be sorted.  
In particular, \cite{cllr-sooscctmp-05} provide an $O(n\log n)$ time algorithm, and since in the worst case computing the contour tree 
requires $\Omega(n\log n)$ time, this algorithm is optimal with respect to the size of the output.  
Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(n\alpha(n))$ 
time\footnote{Where $\alpha(n)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
For all instances the bound for our algorithm either matches the best previous bound, or is faster.
For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
previous algorithms had a running time of $\Omega(n\log n)$

%In particular, the algorithm in \cite{cllr-sooscctmp-05} is worst case optimal with respect to the size of the output.  
%Here we present the first contour tree algorithm which is (near\footnote{Our algorithm is optimal on all instance requiring $\Omega(n\alpha(n))$ time.}) 
%optimal with respect not just to the size of the output, but also its structure.  
%Specifically, the bound for our algorithm either matches \cite{cllr-sooscctmp-05}, or is faster.
%For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
%previous algorithms had a running time of $\Omega(n\log n)$.

This work can be seen as a continuation of the study of instance-optimal geometric algorithms introduced by Afshani \etal \cite{abc-ioga-09}.
In order to prove instance optimality we use a non-trivial combinatorial discharging argument, which we believe is of independent interest.
Of particular note is the use of heavy-light path decompositions, first introduced by Sleator and Tarjan \cite{st-dsdt-83} in their analysis of link-cut trees.

% Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  
%Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
% 
% \medskip\noindent
% Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.

@@

\subsubsection{Worst Case Alternatives}
In worst case analysis an algorithm's running time is judged in terms of its worst 
case behavior over all inputs of a given size.  While there are benefits to such a 
pessimistic view, it fails to capture the whole story.  As such, many alternatives 
have been considered.  For example, one can make assumptions on the distribution of 
inputs, which has led to a number interesting results on average case and probabilistic 
analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
assumptions on the distribution of the input, there is still more to the story.
For example, in the study of parametrized complexity, one takes into account other 
parameters determined by the input other than just its size (for example the maximum 
degree in a graph or whether it is planar).

\paragraph{Output Sensitive Analysis.}
In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
by the input, and so it is natural to also consider how the form of this output structure affects the running time.
One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
significant improvements over their worst case counterparts, their running times are only measured with respect 
to the size of the output and not any structure it may have.

\paragraph{Instance Optimality.}  
Here we are concerned with a more refined analysis than the typical output sensitive analysis.
Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
In order to prove such a result, one considers a family of allowable partitions of the input, 
each of which is defined in terms of the structure of the output.  Each such partition then has 
an associated entropy type cost representing the number of possible instantiations of such a partition.
Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
any algorithm requires the maximum over these costs.

\subsubsection{Contour Trees}

\Ben{Maybe discuss contour tree applications and why they are useful here? Or maybe earlier?}

\paragraph{A History of Contour Trees.}
As discussed above, there has been a long series of improved algorithms for computing contour trees.
Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making 
several passes over the input.

Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, 
where $n$ is the number of mesh vertices.  Their algorithm works in 3 simple stages.  
The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
together the join and split trees to get the contour tree.

There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
of critical points in the input mesh, which could potentially be much smaller than $n$.  In particular, 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 
$3$ dimensional structured meshes.  \Ben{Update distinction between critical and component-critical}
Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
which works for structured or unstructured meshes in any dimension, which is provably optimal with respect 
to the output size but not the output structure, as discussed below.

%As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, 
%one can show that any algorithm to compute the contour tree in the worst case takes 
%$\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree 
%is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a 
%large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only 
%worst case optimal.  In this paper we will consider an output sensitive approach which 
%takes into account the shape of the contour tree.

\begin{figure}[h]\centering
    \includegraphics[width=.27\linewidth]{figs/spike3}%
    \hspace{.35in}
    \includegraphics[width=.27\linewidth]{figs/spike4}%
    \hspace{.4in}
    \includegraphics[width=.22\linewidth]{figs/spikeTree2}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

\paragraph{The Issue of Sorting.}
As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour 
tree by first constructing the join and split trees.  As vertices appear in sorted order 
in any root to leaf path in the join and split trees, and in any maxima to minima path 
in the contour tree, sorting the input vertices is a natural first step.  However, not 
all vertices from the input need to be sorted.  In particular, Chiang \etal 
\cite{cllr-sooscctmp-05} improved the running time by observing that as the join and 
split trees only consist of critical points, it suffices to only sort critical points.  
If a large fraction of the vertices appear in a single path in the join or split tree, 
then paying to sort all the critical points is unavoidable, and therefore in the worst 
case \cite{cllr-sooscctmp-05} is optimal.  However, if the trees are more balanced then 
sorting all critical points together can be very costly.  Specifically, the contour tree 
does not care about the relative ordering of vertices which do not have an 
ancestor-descendant relationship, and so this information does not need to be computed.
See \Fig{relative} for an example.

\begin{figure}[h]\centering
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
    \hspace{.5in}
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
    \caption{On the left, a surface will a balanced contour tree, but whose merge and split trees have long tails.  
    On the right (in clockwise order from the top) the contour, split and merge trees.}
    \label{fig:sorting}
\end{figure}

\paragraph{Suboptimal Join and Split Trees.}
There is also a more fundamental problem with the now standard approach of first computing 
the join and split trees in order to compute the contour tree.  In particular, this approach 
can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  
This is because it is possible to have join and split trees which are long and skinny (i.e. 
costly to compute) whereas the contour tree is balanced.  For example consider the case shown 
in \Fig{sorting}, where the mesh looks like two (thickened) balanced binary trees, one upside 
down, and joined at their roots.  In this case the contour tree is potentially cheap to 
compute as there is no long sorted path.  However, the join and split trees will have long 
sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), 
and therefore require $\Omega(n\log n)$ time to compute.

\subsection{Our Contribution}
\Ben{To be Written}

\paragraph{Organization.}

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/orderGray1}%
    \hspace{.8in}
    \includegraphics[width=.18\linewidth]{figs/orderGray2}
    \caption{On left, downward paint spilling only (each shade of gray represents a different color).  On right, flipping the direction of paint spilling.}
    \label{fig:order}
\end{figure}

@@

\subsection{Shrubs, tall paths, and short paths}

@@

\begin{definition}
 For $p\in P(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
\end{definition}

@@

The following lemma demonstrates that tall paths can ``pay'' for themselves.

\begin{lemma}
\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
% Then, 
% \begin{compactenum}
%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
% \end{compactenum}
\end{lemma}
\begin{proof} \ignore{
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
}
For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
 If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
\end{proof}

\ignore{
For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
Our eventual goal will be to show that by choosing the right path decomposition, this excess 
can be recharged to tall paths. In the next couple sections we introduce the appropriate path decomposition and prove 
some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
}

There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
\emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
short path (i.e. it is a chain of paths adjacent to $p$).

\begin{definition} \label{def:support} A short path $p$ is \emph{supported} if
at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
%
%
% there exists
%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
%and all $q_i$'s are tall.
\end{definition}

Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
given by $\pmax(T)$.
We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
and the remaining nodes are the unsupported short paths.

@@

We split the summation into tall, short, and unsupported short paths.
\begin{eqnarray*} 
\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
\end{eqnarray*}
The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
The following claim shows that this in turn is at most the last term.

@@

\begin{proof} 
Pick $p \in \cL'$. 
%Let the first $|H_p|/200$ vertices the lie in tall paths be denoted as $H'_p$. 
%Observe that for any $v \in H'_p$, 
%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in tall paths,
%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%
%
As we traverse the tall support chain of $p$, there
are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
Since all the vertices lie in tall paths,
we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.

Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.

@@

The first term (unsupported short paths) can be charged to the second term (supported short paths).
This is where the critical \Lem{cu-root} plays a role.

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

\ignore{
Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of short paths, 
and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
is short. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.

\begin{lemma} \label{lem:short} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{lemma}

@@

\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into tall and short paths.
We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
For the first term, we apply \Lem{short} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
\end{proof}
